\chapter{The Physical Environment}
\label{cha:environment}

\section{Overview}

Wireless networks are heavily affected by the physical environment, and the
requirements for today's ubiquitous wireless communication devices are
increasingly demanding. Cellular networks serve densely populated urban
areas, wireless LANs need to be able to cover large buildings with several
offices, low-power wireless sensors must tolerate noisy industrial
environments, batteries need to remain operational under various external
conditions, and so on.

The propagation of radio signals, the movement of communicating agents, battery
exhaustion, etc., depend on the surrounding physical environment. For example,
signals can be absorbed by objects, can pass through objects, can be refracted
by surfaces, can be reflected from surfaces, battery nominal capacity might depend on
external temperature, etc. These effects cannot be ignored in high-fidelity
simulations.

In order to help the modeling process, the model of the physical
environment is separated from the rest of the simulation models. The main
goal of the physical environment model is to describe buildings, walls,
vegetation, terrain, weather, and other physical objects and conditions
that might have effects on radio signal propagation, movement, batteries,
etc. This separation makes the model reusable by all other simulation
models that depend on these circumstances.

The following sections provide a brief overview of the physical environment
model.

\section{Physical Objects}

The most important aspect of the physical environment is the objects which
are present in it, as they are obstructing radio signals. For example,
simulating an indoor Wifi scenario needs to model walls, floors, ceilings,
doors, windows, furniture, etc. These objects are located in space, have
shapes and materials. The INET physical layer infrastructure supports basic
shapes and homogeneous materials, which simplifies description and still
allows for a reasonable approximation of reality. Physical objects in INET
have the following properties:

\begin{itemize}
  \item \textit{shape}: describes the 3-dimensional shape of the object,
    independent of its position and orientation
  \item \textit{position}: determines where the object is located in the 3-dimensional space
  \item \textit{orientation}: determines how the object is rotated relative to its
    default orientation
  \item \textit{material}: describes material-specific physical properties
  \item \textit{graphical properties}: provides parameters for better visualization
\end{itemize}

Physical objects in INET are stationary, they cannot change their position
or orientation over time.

Since the shape of the physical objects might be quite diverse, the model
is designed to be extensible with new shapes. Concave shapes are not yet
supported, such shapes can be represented by splitting them up into smaller,
convex parts. The current implementation provides the following shapes:

\begin{itemize}
  \item \cppclass{Sphere}: specified by a radius
  \item \cppclass{Cuboid}: specified by a length, a width, and a height
  \item \cppclass{Prism}: specified by a 2-dimensional polygon base and a height
  \item \cppclass{Polyhedron}: specified by the convex hull of a set of
    3-dimensional vertices
\end{itemize}

In order to model the physical environment in detail, a scenario might contain
several thousands or even more physical objects. Simulation models might
need to query these objects quite often. For example, when the physical layer
computes obstacle loss for a transmission, it needs to find the obstructing
physical objects for each receiver. This requires computing the intersection
between physical objects and the path traveled by the radio signal.

To speed up the computation of intersections, INET stores physical objects
in a highly efficient data structure, which currently can be one of the
following:

\begin{itemize}
  \item \nedtype{GridObjectCache}: organizes objects into a 3D spatial grid with
    a configurable constant cell size, where cells contain the objects that
    intersect with them
  \item \nedtype{BVHObjectCache}: organizes objects into a 3D tree, where
    leaves contain a configurable number of closely positioned objects.
    (This data structure is similar to quadtree and octree, but is designed for
    storing finite-sized objects.)
\end{itemize}

The physical environment uses \nedtype{GridObjectCache} by default.

\section{Global Physical Properties}

The INET physical environment infrastructure stores the following global
properties:

\begin{itemize}
  \item \textit{space limits}: global bounds for the 3-dimensional space
  \item \textit{ambient temperature}: global parameter for temperature-dependent models
\end{itemize}

Space limits are useful for limiting the propagation and reflection of
radio signals, to constrain movement of communicating agents, and for
detecting positioned physical objects.

Ambient temperature can be useful for modeling batteries, as it affects the
maximum capacity, internal resistance, self-discharge and other properties
of real-life electrochemical energy storage devices.

\section{Initialization}

In order to easily define thousands of physical objects the model needs to use a
flexible and concise representation. For this purpose, the physical environment
uses an XML file format. This file can be used to define physical objects along
with shapes, materials used by them. The following example demonstrates the
syntax of this XML file format:

\begin{verbatim}
<environment>
  <!-- shapes and materials -->
  <shape id="1" type="sphere" radius="10"/>
  <shape id="2" type="cuboid" size="20 30 40"/>
  <shape id="3" type="prism" height="100"
         points="0 0 100 0 100 100 0 100"/>
  <shape id="4" type="polyhedron"
         points="0 0 0 100 0 0 100 100 0 0 100 0 ..."/>
  <material id="1" resistivity="100"
            relativePermittivity="4.5" relativePermeability="1"/>
  <!-- objects that use previously defined shapes and materials -->
  <object position="min 100 200 0" orientation="45 -30 0"
          shape="1" material="1"
          line-color="0 0 0" fill-color="112 128 144" opacity="0.5"/>
  <object position="min 100 200 0" orientation="45 -30 0"
          shape="cuboid 20 30 40" material="concrete"
          line-color="0 0 0" fill-color="112 128 144" tags="Building"/>
</environment>
\end{verbatim}

For more details on the XML file format, please refer to the documentation in the
\nedtype{PhysicalEnvironment} NED file.

\section{Visualization}

Understanding the abstract physical environment model from numerical log output,
or debugging simulation models based on the numerical data structures is
difficult. Therefore the physical environment supports visualizing the physical
objects on the graphical user interface. The visualization is done by drawing
the physical objects on the parent compound module as a separate layer. In order
to help to distinguish among physical objects they have the following graphical
properties:

\begin{itemize}
  \item \textit{line width}: surface outline
  \item \textit{line color}: surface outline
  \item \textit{fill color}: surface filling
  \item \textit{opacity}: surface outline and filling
  \item \textit{tags}: allows filtering objects on the graphical user interface
\end{itemize}

Although the physical objects are really modeled in 3 dimensions, the
visualization is only 2 dimensional. The projection which converts physical
objects to 2 dimensional shapes can be parameterized with an arbitrary view
angle. The default view angle is the Z axis, in which case the layer shows the
physical objects from above. The view angle can be changed during runtime using
the parameter editor of the graphical user interface.

The projection is also used by other models for their own visualizations. This
approach makes visualizations well integrated with each other. For example, the
mobility models position network nodes, the physical layer draws ongoing
transmissions, the obstacle loss draws intersections between transmissions and
objects, etc., on the canvas according to the projection described above.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "usman"
%%% End:

